\bf Model Visualization Techniques for a Social Network Model

### 2017-04-17 paper.Rnw
up to section 3

Network data is of frequent interest to researchers in a wide array of fields. There are techonological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, e.g. neural networks, and social networks, just to name a few \citep{newman}. Each of these examples have one thing in common: their data structure. There are always units of observation, which we refer to throughout as nodes, and connections of some kind between those units, which we will call edges. The networks can also vary over time, for example when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network. The nodes and edges themselves also have inherent variables of interest, such as the the affiliation of authors in a coauthorship network, and the number of times two authors have collaborated.

The multiple layers, so to speak, of network data structures create unique struggles to network analysts. Some questions that researchers aim to answer can be: How does the strength of a connection between two nodes affect the overall structure of the network? Do node-level differences affect the formation or destruction of edges? Which views of the data are most informative for communicating effects and results of statistical analyses? These are, of course, just a few broad questions, and we will focus on only the latter.

@@

There are two primary methods used to visualize networks: node-link diagrams and matrix visualization. We will use the toy data set show below to demonstrate these two methods.

<<toyex>>=
set.seed(81393)
nodes <- data.frame(id = 1:5, group = sample(LETTERS[1:2], size = 5, replace = T), stringsAsFactors = F)
g <- igraph::as_data_frame(igraph::random.graph.game(n = 5, p = .4))
toynet <- list(nodes = nodes, edges = rbind(g, c(5,4)))
toynet
@

The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes. But, because there is no natural placement of the points, a random placement is used, then adjusted iteratively via a layout algorithm, of which there are many. The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.

Some layout algorithms were designed to mimic physical systems, drawing the graphs based on the "forces" connecting them. The network's edges act as springs pushing and pulling the nodes in 2D space. Others use properties of the network's adjaceny matrix in order to place the nodes. Some commonly used layout algorithms that treat networks like physical systems are Kamada-Kawai \citep{kamadakawai}, Fructerman-Reingold  \citep{fruchterman-reingold}. Other algorithms inclue multi-dimensional scaling, which relies on a defined distance matrix, or an eigen value, which depends on the eigenvalues of the network's adjacency matrix.

The second primary method for network visualization uses the adjacency matrix of the network. An example of this type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example. This method is particularly useful when the network is very complex, dense, or large. Experimental studies have shown this to be true. For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs. This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, but a node-link diagram may underrepresent the edge count with overlapping edges.

>>>>

Network data is of frequent interest to researchers in a wide array of fields. There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, e.g. neural networks, and social networks, just to name a few \citep{newman}. Each of these examples have one thing in common: their data structure. There are always units of observation, which we refer to throughout this paper as {\emph nodes}, and connections of some kind between those units, which we will call {\emph edges}. Networks might change over time, for example when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network. Nodes and edges themselves also have inherent variables of interest, such as the  affiliation of authors in a co-authorship network, or the number of times two authors have collaborated.

The multiple layers of network data structures create unique struggles to network analysts. Some questions that researchers aim to answer can be: How does the strength of a connection between two nodes affect the overall structure of the network? Do node-level differences affect the formation or destruction of edges? Which views of the data are most informative for communicating effects and results of statistical analyses? These are, of course, just a few broad questions. In this paper we want to focus on the visual exploration of network data and models.

@@

There are two primary methods used to visualize networks: node-link diagrams and matrix visualization.
%\hh{As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by some edges: XXX list edges as $1 \mapto 2, 2 \mapto 4, ...$ }
We will use the toy data set shown below to demonstrate these two methods.

\hh{XXX describe the dataset in words, but not with the code. }
<<toyex>>=
set.seed(81393)
nodes <- data.frame(id = 1:5, group = sample(LETTERS[1:2], size = 5, replace = T), stringsAsFactors = F)
g <- igraph::as_data_frame(igraph::random.graph.game(n = 5, p = .4))
toynet <- list(nodes = nodes, edges = rbind(g, c(5,4)))
toynet
@

The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes. But because there is no natural placement of the points, a random placement is used, then adjusted iteratively via a layout algorithm, of which there are many. The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.

Some layout algorithms were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. The network's edges act as springs pushing and pulling the nodes in a lower dimensional (usually two-dimensional) space. %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes. \hh{XXX how does the adjacency matrix feature here? The adjacency matrix is  a different mathematical representation of a network, but describes the same network as a node-link diagram.} 
Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, treat networks like physical systems. Other algorithms are based on multi-dimensional scaling, which relies on a defined distance matrix, or an eigen value, which depends on the eigenvalues of the network's adjacency matrix.

The second primary method for network visualization uses the adjacency matrix of the network. \hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} An example of this type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example. This method is particularly useful when the network is very complex, dense, or large. Experimental studies have shown this to be true. For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs. \hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented? Why would directed graphs make that better? wouldn't that just add to the number of edges? } This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, but a node-link diagram may underrepresent the edge count with overlapping edges.
